Off the beaten path: \\ Gaia reveals GD-1 stars outside of the main stream

### 2018-05-01 paper/gd1.tex
further trims

\begin{abstract}\noindent % trust me
Using the exquisite data from the \gaia\ second data release combined with
\pans\ photometry,  we present a sample of highly-probable members of the
longest cold stream in the Milky Way -- GD-1.
The resulting map of GD-1: (1) extends the apparent length of the stream by
$20^\circ$, (2) reveals a plausible location for the progenitor, (3) detects 
significant density variations along the stream, and (4) indicates
the existence of stream members off the main stream track.
% Tidally-disrupted globular clusters are transformed into thin, dynamically-cold
% streams of stars that are extremely valuable tracers of the large- and
% small-scale distribution of mass in the Galaxy.
% Most stellar streams discovered around the Milky Way reside in the Galactic halo
% and are therefore primarily sensitive to the dark matter distribution.
% The longest and most prominent thin stream in the Milky Way --- the GD-1 stream --- has been used
% to constrain the shape of the inner halo, and is a promising place to search for
% signatures of dark matter subhalo interactions.
% selected to have
% small parallaxes and retrograde proper motions using data from the \gaia\
% mission, and \pans\ photometry consistent with an old and metal-poor population
% at a distance of $\sim10~\kpc$.
% This selection extends the stream by $20^\circ$ in apparent length, reveals a
% plausible location of the progenitor, and shows significant density variations
% along the stream in unprecedented detail.
% This new view of the stream reveals significant density variations
% along the stream in unprecedented detail.
% In addition to confirming several prominent gaps detected in photometric maps of
% GD-1, for the first time we find evidence of stream members off of the main
% track.
These features could be an indication of complex internal kinematics within the
progenitor unlike any of the known globular clusters, or a result of a stream
interaction with a massive perturber.
% Detailed modeling of the positional and velocity offsets
% or an indication of complex internal kinematics within the
% stream progenitor.
% Velocity offset between these features and the stream is an order of magnitude larger than the variance in velocity fields of known globular cluster, thus favoring the encounter origin.
% If this is confirmed, off-stream features will enable a robust estimate of the perturber's mass.
% If caused by a subhalo interaction, we estimate that the detected underdensity
% and the associated spur could have been induced by a perturbation of a
% $\sim10^7\,\rm M_\odot$(?) object $\sim 0.5\,$Gyr ago.
\end{abstract}

@@

Its length and location in the Galactic halo (pericentric distance
$r_\textrm{peri} \sim 13~\kpc$) make GD-1 an ideal object for constraining dark
matter properties in the inner Milky Way.
Both from fitting orbits to binned phase-space measurements along the GD-1
stream (\citealt{Koposov:2010}) and from modeling the stream track in
phase-space coordinates \citep{Bowden:2015, Bovy:2016}, it appears that the dark
matter distribution within $\approx20\,\rm kpc$ is close to spherical.
This is consistent with findings from simulated Milky Way-like galaxies:
simulated dark-matter-only halos tend to be triaxial \citep{Allgood:2006}, but
baryonic processes tend to make their inner regions more spherical
\citep{Zhu:2016}.

Its implied old dynamical age, combined with its orbital properties, makes GD-1
a prime stream to also study the small scale structure of dark matter.
While density variations in streams can be induced from interactions with giant
molecular clouds \citep{Amorisco:2016} or resonant encounters with the bar
\citep{Pearson:2017}, these baryonic effects are unlikely to affect GD-1 because
of its large pericentric distance and retrograde orbit with respect to the
Galactic bar.
Therefore, observed density variations in the GD-1 stream could instead be an
indication of past interactions with dark matter subhalos
\citep[e.g.,][]{Ngan:2014}.
By estimating a stream age of $\approx 4~\textrm{Gyr}$, assuming a progenitor
mass of $10^5~\msun$, and by assuming properties of the subhalo population from
% dark matter-only simulations \citep[scaled to the Milky Way halo mass and
% reduced in number by a factor of 3 to account for disruption by the Galactic
% disk]{Springel:2008, Diemand:2008}, \citet{Erkal:2016} predict that GD-1 could
dark matter-only simulations \citep{Springel:2008, Diemand:2008},
\citet{Erkal:2016} predict that GD-1 could have up to one significant, wide
($\approx 5$--$7^\circ$) gap caused by an interaction with a
$10^6$--$10^7~\msun$ subhalo.
A definitive ruling on the presence of dark matter subhalos in this mass regime
would directly inform the nature of dark matter \citep[e.g.,][]{Bullock:2017}.

@@

We retrieve \gaia\ data along the previously-identified track of the GD-1 stream
from the \gaia\ science archive\footnote{\url{https://gea.esac.esa.int/}} by
selecting sources with small parallax, $\varpi < 1~\textrm{mas}$, and select
batches of stars that have small latitude in the heliocentric GD-1 coordinate
system, $(\phi_1, \phi_2)$, defined in \cite{Koposov:2010}.
We convert the equatorial sky coordinates and proper motions %of the parallax-selected stars
to the GD-1 stream system, and correct the proper motions for solar reflex
motion by assuming that stars at a given stream longitude, $\phi_1$, have a
distance given by $d(\phi_1) = (0.05 \, \phi_1 + 10)~\textrm{kpc}$;
we assume a solar velocity $\bs{v}_\odot = (11.1, 232.24, 7.25)~\kms$
(\citealt{Schonrich:2010, Bovy:2015}).
\figurename~\ref{fig:selection} (top right) shows stars with $|\phi_2| <
1^\circ$ in solar-reflex-corrected proper motion components in the GD-1 system,
$\mu_{\phi_1, \odot}$ and $\mu_{\phi_2, \odot}$.
GD-1 members are visible as the overdensity around $(\mu_{\phi_1, \odot},
\mu_{\phi_2, \odot}) \approx (-8, 0)~\masyr$.

As a first selection of GD-1 member stars, we select stars with proper motions
$-9 < \mu_{\phi_1, \odot} < -4.5~\masyr$ and $-1.7 < \mu_{\phi_2, \odot} <
1~\masyr$, as visualized by the rectangle in the top right panel of
\figurename~\ref{fig:selection}.
\figurename~\ref{fig:selection} (top left) shows all stars that pass the
selection, plotted in the GD-1 coordinate system.
The stream is already identifiable as the overdensity of stars around $\phi_2 =
0$ between longitudes $-60^\circ \lesssim \phi_1 \lesssim 0^\circ$.

@@

To improve the contrast of the stream over the background, we cross-match the
sample to the \pans\ photometric catalog. % and use the photometry to clean the sample.
\figurename~\ref{fig:selection} (bottom right) shows the proper-motion-selected
candidate stream stars within $\left|\phi_2\right| < 1^\circ$ in a \pans\
color-magnitude diagram (CMD) de-reddened following \citet{Schlafly:2011}.
GD-1 stellar population stands out from the Milky Way foreground, and is well matched by a 12\,Gyr, $\feh = -1.35$ MIST isochrone at a distance of $7.8\,\kpc$ \citep[red line in Figure~\ref{fig:selection}][]{Dotter:2016, Choi:2016, Paxton:2011}. 
% The over-plotted isochrone (red line) is from the MESA Isochrones \& Stellar
% Tracks (MIST; \citealt{Dotter:2016, Choi:2016, Paxton:2011}) and represents a
% $12~\textrm{Gyr}$ old population with $\feh = -1.35$ at a distance of
% $7.8~\kpc$.
We use this isochrone to generate a polygonal selection in de-reddened $g-i$
color and apparent $g$-band magnitude (shaded region in the
bottom right panel of \figurename~\ref{fig:selection}).

@@

To compute initial conditions for the stream model, we fit an orbit to the
observed properties of the stream in a fixed Milky Way model consisting of a
disk (\citealt{Miyamoto:1975}), a bulge (\citealt{Hernquist:1990}), and a
spherical dark matter halo (\citealt{Navarro:1996}).
We set the disk scale-length parameters to match the disk model found in
\citet{Bovy:2015}; other parameters are set to produce an approximately flat
disk rotation curve with a circular velocity at the solar circle,
$v_{\textrm{circ}, \odot} \approx 230~\kms$.
%use L-BFGS-B optimization as implemented in the \package{scipy}
% \texttt{Python} package to
We maximize the likelihood of the orbit given sky track and proper motions from
this work, with mean distance and radial velocities from
\cite{Koposov:2010}.
% We do not vary the parameters of the Milky Way model in the fit because we are
% only interested in producing a comparison stream model. % of the stream properties.

@@

\begin{figure}
\begin{center}
\includegraphics[width=\textwidth]{features.pdf}
\end{center}
\caption{
Color-magnitude and proper-motion diagrams of fields associated with a spur above GD-1 (top), and fields related to the blob below GD-1 (bottom).
The top and bottom left panels show properties in stream fields (darkest rectangles in middle panel), both middle panels present spur and blob (medium rectangles), while the right-most top and bottom panels are selected opposite from spur and blob (lightest rectangles).
Spatial selection boxes are shown in the middle panel as rectangles, and in all CMD and proper-motion panels as shaded regions.
Both the spur and the blob fields have more stars in the CMD selection box than the corresponding control fields.
Proper motions of these stars follow the distribution of stars in adjacent stream fields, thus confirming the association of these off-track features with GD-1.
}
\label{fig:features}
\end{figure}

>>>>

\begin{abstract}\noindent % trust me
Tidally-disrupted globular clusters are transformed into thin, dynamically-cold
streams of stars that are extremely valuable tracers of the large- and
small-scale distribution of mass in the Galaxy.
Using the exquisite data from the \gaia\ second data release combined with
\pans\ photometry, we present a sample of highly-probable members of the
longest cold stream in the Milky Way -- GD-1.
The resulting map of GD-1: (1) extends the apparent length of the stream by
$20^\circ$, (2) reveals a plausible location for the progenitor, (3) detects 
significant density variations along the stream, and (4) indicates
the existence of stream members off the main stream track.
The additional length and proper treatment of the progenitor will aid in dynamical modeling of GD-1 and map the Galaxy on large scales, while its complex morphology points to a turbulent history, potentially informing on massive, small-scale perturbers within 30\,kpc.

% Most stellar streams discovered around the Milky Way reside in the Galactic halo
% and are therefore primarily sensitive to the dark matter distribution.
% The longest and most prominent thin stream in the Milky Way --- the GD-1 stream --- has been used
% to constrain the shape of the inner halo, and is a promising place to search for
% signatures of dark matter subhalo interactions.
% selected to have
% small parallaxes and retrograde proper motions using data from the \gaia\
% mission, and \pans\ photometry consistent with an old and metal-poor population
% at a distance of $\sim10~\kpc$.
% This selection extends the stream by $20^\circ$ in apparent length, reveals a
% plausible location of the progenitor, and shows significant density variations
% along the stream in unprecedented detail.
% This new view of the stream reveals significant density variations
% along the stream in unprecedented detail.
% In addition to confirming several prominent gaps detected in photometric maps of
% GD-1, for the first time we find evidence of stream members off of the main
% track.
These features could be an indication of complex internal kinematics within the
progenitor unlike any of the known globular clusters, or a result of a stream
interaction with a massive perturber.
% Detailed modeling of the positional and velocity offsets
% or an indication of complex internal kinematics within the
% stream progenitor.
% Velocity offset between these features and the stream is an order of magnitude larger than the variance in velocity fields of known globular cluster, thus favoring the encounter origin.
% If this is confirmed, off-stream features will enable a robust estimate of the perturber's mass.
% If caused by a subhalo interaction, we estimate that the detected underdensity
% and the associated spur could have been induced by a perturbation of a
% $\sim10^7\,\rm M_\odot$(?) object $\sim 0.5\,$Gyr ago.
\end{abstract}

@@

Its length and location in the Galactic halo (pericentric distance
$r_\textrm{peri} \sim 13~\kpc$) make GD-1 an ideal object for constraining dark
matter properties in the inner Milky Way.
Both from fitting orbits to binned phase-space measurements along the GD-1
stream (\citealt{Koposov:2010}) and from modeling the stream track in
phase-space coordinates \citep{Bowden:2015, Bovy:2016}, it appears that the dark
matter distribution within $\approx20\,\rm kpc$ is close to spherical, consistent 
with findings from simulated Milky Way-like galaxies \citep{Zhu:2016}.
% This is consistent with findings from simulated Milky Way-like galaxies:
% simulated dark-matter-only halos tend to be triaxial \citep{Allgood:2006}, but
% baryonic processes tend to make their inner regions more spherical
% \citep{Zhu:2016}.

Its implied old dynamical age, combined with its orbital properties, makes GD-1
a prime stream to also study the small scale structure of dark matter.
While density variations in streams can be induced from interactions with giant
molecular clouds \citep{Amorisco:2016} or resonant encounters with the bar
\citep{Pearson:2017}, these baryonic effects are unlikely to affect GD-1 because
of its large pericentric distance and retrograde orbit with respect to the
Galactic bar.
Therefore, observed density variations in the GD-1 stream could instead be an
indication of past interactions with dark matter subhalos
\citep[e.g.,][]{Ngan:2014}.
% By estimating a stream age of $\approx 4~\textrm{Gyr}$, assuming a progenitor
% mass of $10^5~\msun$, and by assuming properties of the subhalo population from
% dark matter-only simulations \citep{Springel:2008, Diemand:2008},
\citet{Erkal:2016} predict that a 4\,Gyr old GD-1-like stream in a Milky Way-like galaxy
could have up to one significant, wide
($\approx 5$--$7^\circ$) gap caused by an interaction with a
$10^6$--$10^7~\msun$ subhalo.
A definitive ruling on the presence of dark matter subhalos in this mass regime
would directly inform the nature of dark matter \citep[e.g.,][]{Bullock:2017}.
% dark matter-only simulations \citep[scaled to the Milky Way halo mass and
% reduced in number by a factor of 3 to account for disruption by the Galactic
% disk]{Springel:2008, Diemand:2008}, \citet{Erkal:2016} predict that GD-1 could

@@

We retrieve \gaia\ data along the previously-identified track of the GD-1 stream
from the \gaia\ science archive\footnote{\url{https://gea.esac.esa.int/}} by
selecting sources with small parallax, $\varpi < 1~\textrm{mas}$ in fields along the GD-1 stream
% , and select
% batches of stars that have small latitude in the heliocentric GD-1 coordinate
% system, $(\phi_1, \phi_2)$, defined in \cite{Koposov:2010}.
We convert the equatorial sky coordinates and proper motions %of the parallax-selected stars
to the GD-1 coordinate system \citep[$\phi_1, \phi_2$,][]{Koposov:2010}, and correct the proper motions for solar reflex
motion by assuming that stars at a given stream longitude, $\phi_1$, have a
distance given by $d(\phi_1) = (0.05 \, \phi_1 + 10)~\textrm{kpc}$;
we assume a solar velocity $\bs{v}_\odot = (11.1, 232.24, 7.25)~\kms$
(\citealt{Schonrich:2010, Bovy:2015}).
\figurename~\ref{fig:selection} (top right) shows stars with $|\phi_2| <
1^\circ$ in solar-reflex-corrected proper motion components in the GD-1 system,
$\mu_{\phi_1, \odot}$ and $\mu_{\phi_2, \odot}$.
GD-1 members are visible as the overdensity around $(\mu_{\phi_1, \odot},
\mu_{\phi_2, \odot}) \approx (-8, 0)~\masyr$.

As a first selection of GD-1 member stars, we choose stars with proper motions
$-9 < \mu_{\phi_1, \odot} < -4.5~\masyr$ and $-1.7 < \mu_{\phi_2, \odot} <
1~\masyr$ (rectangle in the top right panel of
\figurename~\ref{fig:selection}).
\figurename~\ref{fig:selection} (top left) shows all stars that pass the
selection, plotted in the GD-1 coordinate system.
The stream is already identifiable as the overdensity of stars around $\phi_2 =
0$ between longitudes $-60^\circ \lesssim \phi_1 \lesssim 0^\circ$.

@@

To improve the contrast of the stream over the background, we cross-match the
sample to the \pans\ photometric catalog. % and use the photometry to clean the sample.
\figurename~\ref{fig:selection} (bottom right) shows the proper-motion-selected
candidate stream stars within $\left|\phi_2\right| < 1^\circ$ in a \pans\
color-magnitude diagram (CMD) de-reddened following \citet{Schlafly:2011}.
GD-1 stellar population stands out from the Milky Way foreground, and is well matched by a 12\,Gyr, $\feh = -1.35$ MIST isochrone at a distance of $7.8\,\kpc$ \citep[red line in Figure~\ref{fig:selection},][]{Dotter:2016, Choi:2016, Paxton:2011}. 
% The over-plotted isochrone (red line) is from the MESA Isochrones \& Stellar
% Tracks (MIST; \citealt{Dotter:2016, Choi:2016, Paxton:2011}) and represents a
% $12~\textrm{Gyr}$ old population with $\feh = -1.35$ at a distance of
% $7.8~\kpc$.
We use this isochrone to generate a polygonal selection in de-reddened $g-i$
color and apparent $g$-band magnitude (shaded region in the
bottom right panel of \figurename~\ref{fig:selection}).

@@

To compute initial conditions for the stream model, we fit an orbit to the
observed properties of the stream in a fixed Milky Way model similar to \citet{Bovy:2015}.
% consisting of a
% disk (\citealt{Miyamoto:1975}), a bulge (\citealt{Hernquist:1990}), and a
% spherical dark matter halo (\citealt{Navarro:1996}).
% We set the disk scale-length parameters to match the disk model found in
% \citet{Bovy:2015}; other parameters are set to produce an approximately flat
% disk rotation curve with a circular velocity at the solar circle,
% $v_{\textrm{circ}, \odot} \approx 230~\kms$.
%use L-BFGS-B optimization as implemented in the \package{scipy}
% \texttt{Python} package to
We maximize the likelihood of the orbit given sky track and proper motions from
this work, with mean distance and radial velocities from
\cite{Koposov:2010}.
% We do not vary the parameters of the Milky Way model in the fit because we are
% only interested in producing a comparison stream model. % of the stream properties.

@@

\begin{figure}
\begin{center}
\includegraphics[width=\textwidth]{features.pdf}
\end{center}
\caption{
Color-magnitude and proper-motion diagrams of fields associated with a spur above GD-1 (top), and fields related to the blob below GD-1 (bottom).
The top and bottom left panels show properties in stream fields (darkest rectangles in middle panel), both middle panels present spur and blob (medium rectangles), while the right-most top and bottom panels are selected opposite from spur and blob (lightest rectangles).
Selection boxes are shown in all CMD and proper-motion panels as shaded regions.
Both the spur and the blob fields have more stars in the CMD selection box than the corresponding control fields.
Proper motions of these stars follow the distribution of stars in adjacent stream fields, thus confirming the association of these off-track features with GD-1.
}
\label{fig:features}
\end{figure}
